A space-time skew-$t$ model for threshold exceedances

### 2015-11-04 LaTeX/student-competition/skewnormal.tex
updates for student competition

% \documentclass[11pt]{article}
\documentclass[useAMS,usenatbib,referee]{biom}
% \usepackage{amssymb, amsthm, amsmath}
\usepackage{amssymb, amsmath}
\usepackage{bm}
\usepackage{graphicx}
% \usepackage[authoryear]{natbib}
\usepackage{natbib}
\usepackage{bm}
\usepackage{verbatim}
\usepackage{lineno}
\usepackage{times}
\usepackage{soul}
\usepackage{color}
\usepackage{enumitem}
% \usepackage{enumerate}
\usepackage{setspace}

@@

% \usepackage[left=1in,top=1in,right=1in]{geometry}
% \pdfpageheight 11in
% \pdfpagewidth 8.5in
% \linespread{2.0}
\input{mycommands.sty}

\title[A Space-time Skew-$t$ Model for Threshold Exceedances]{A space-time skew-$t$ model for threshold exceedances}
\author
{Samuel A Morris$^{1,*}$\email{samorris@ncsu.edu},
Brian J Reich$^{1}$,
Emeric Thibaud$^{2}$, and
Daniel Cooley$^{2}$\\
$^{1}$Department of Statistics, North Carolina State University, Raleigh, North Carolina, U.S.A. \\
$^{2}$Department of Statistics, Colorado State University, Fort Collins, Colorado, U.S.A.}

\begin{document} %\linenumbers
% \begin{center}
% {\Large {\bf A space-time skew-$t$ model for threshold exceedances}}\\

% {\large Samuel A Morris\footnote[1]{North Carolina State University}, Brian J Reich\footnotemark[1]{}, Emeric Thibaud\footnote[2]{Colorado State University}, and Daniel Cooley\footnotemark[2]{}}

% \today
% \end{center}

\pagerange{\pageref{firstpage}--\pageref{lastpage}}

@@

\begin{abstract}
To assess the compliance of air quality regulations, the Environmental Protection Agency (EPA) must know if a site exceeds a pre-specified threshold.
In the case of ozone, the threshold for compliance is fixed at 75 parts per billion, which is high, but not extreme at all locations.
We present a new method based on the spatial skew-$t$ process.
Our method incorporates a random partition to permit long-distance asymptotic independence while allowing for sites that are near one another to be asymptotically dependent, and we incorporate thresholding to allow the tails of the data to speak for themselves.
We also introduce a transformed AR(1) time-series to allow for temporal dependence.
Finally, our model allows for high-dimensional Bayesian inference that is comparable in speed to traditional geostatistical methods for large datasets.
We apply our method to an ozone analysis for July 2005, and find that our model improves over both Gaussian and max-stable methods in terms of predicting exceedances over a fixed threshold.
\end{abstract}

\begin{keywords}
Skew-$t$, random partition, MCMC, extreme value analysis, spatio-temporal
\end{keywords}

\maketitle

@@

\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=0.75\linewidth]{plots/ozone-10jul-us.pdf}
  \caption{Ozone values (ppb) on July 10, 2005}
  \label{fig:ozone-10jul}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.5\linewidth]{plots/chi-h.pdf}
  \caption{Extremal dependence measure $\chi(h)$, as a function of distance, $h$, for $K = 1, 3, 5$, and $10$ knots.}
  \label{fig:chi}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{plots/bsplots-mean.pdf}
  \caption{Brier scores relative to the Gaussian method for simulation study results. A ratio lower than 1 indicates that the method outperforms the Gaussian method.}
  \label{fig:simbrierscores}
\end{figure}

\section{Data analysis}\label{s:analysis}
We consider daily observations of maximum 8-hour ozone measurements for the 31 days of July 2005 at 1,089 Air Quality System (AQS) monitoring sites in the United States as the response (see Figure \ref{fig:ozone-10jul}).
For each site, we also have covariate information containing the estimated ozone from the Community Multi-scale Air Quality (CMAQ) modeling system.
Initially, we fit a linear regression assuming a mean function of
\begin{align}
  \text{E}[Y_i(\bs)] = \beta_0 + \beta_1 \cdot \text{CMAQ}_t(\bs). \label{eq:datamean}
\end{align}
Figure \ref{fig:ozone-qq} shows a Q-Q plot of the residuals compared to a skew-$t$ distribution with 10 d.f. and $\alpha = 1$.
\begin{figure}
  \centering
  \includegraphics[width=0.5\linewidth]{plots/qq-res.pdf}
  \caption{Q-Q plot of the residuals for a skew-$t$ distribution with 10 d.f. and $\alpha = 1$ (right)}
  \label{fig:ozone-qq}
\end{figure}

Standard exploratory data analysis techniques for extremal dependence are very challenging with only 31 days worth of data because it is difficult to estimate extreme quantiles at each site to obtain empirical estimates of $\chi$.
Despite the fact that there is only one month of data, we can get some sense of extremal dependence between sites by looking at joint occurrences of high sample quantiles.
For example, Figure \ref{fig:bivariateozone} suggests there is more agreement between sites that are close to one another than sites that are far from one another.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{plots/daily-quantiles-ozone.pdf}
  \caption{Daily quantiles for two monitoring locations near Columbus, OH (left) and daily quantiles for a monitoring location in Los Angeles, CA and Columbus, OH (right)}
  \label{fig:bivariateozone}
\end{figure}
Another aspect that distinguishes our approach from more traditional extremes analyses, is how the threshold is selected.
In our example, a threshold of 75 ppb which corresponds to $q(92)$ for all observations, but marginally it represents anywhere from $q(0.06)$ to $q(1)$.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{plots/bs-ozone.pdf}
  \caption{Relative Brier scores for time-series models (left) and non-time-series models (right). Relative brier score for the max-stable model is between 1.07 and 1.15}
  \label{fig:bs-ozone}
\end{figure}
\begin{table}
  \small
  \caption{Top two performing models for ozone analysis at extreme quantiles with Relative Brier score}
  \label{tbl:ozoneresults}
  \centering
  \begin{tabular}{|l|l l l c|l l l c|}
    \cline{2-9}
    \multicolumn{1}{c|}{}  & \multicolumn{4}{c|}{1st} & \multicolumn{4}{c|}{2nd} \\
    \hline
    $q(0.90)$  & No time series & $K=7$  & $T=0$  & BS: 0.980 &
                 No time series & $K=9$  & $T=0$  & BS: 0.980 \\
    $q(0.95)$  & No time series & $K=15$ & $T=50$ & BS: 0.970 &
                 No time series & $K=9$  & $T=50$ & BS: 0.970\\
    $q(0.98)$  & No time series & $K=5$  & $T=50$ & BS: 0.945 &
                 No time series & $K=10$ & $T=50$ & BS: 0.946\\
    $q(0.99)$  & Time series    & $K=10$ & $T=75$ & BS: 0.912 &
                 Time series    & $K=6$  & $T=75$ & BS: 0.913\\
    $q(0.995)$ & Time series    & $K=6$  & $T=75$ & BS: 0.917 &
                 Time series    & $K=10$ & $T=75$ & BS: 0.918\\
    \hline
  \end{tabular}
\end{table}

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{plots/q99-ozone.pdf}
  \caption{Panels (a) -- (d) give the posterior predictive $\widehat{q}(0.99)$ for the month of July under four different models, panel (e) gives the difference between $\widehat{q}(0.99)$ in panels (d) and (a), panel (f) gives the difference between $\widehat{q}(0.99)$ in panels (d) and (b).}
  \label{fig:ozoneq99}
\end{figure}

@@

\backmatter

@@

\section*{Supplementary Materials}
Web Appendices A--E are available with this paper at the Biometrics website on Wiley Online Library. \vspace*{-8pt}

\bibliographystyle{biom}
\bibliography{library}

>>>>

\documentclass[11pt]{article}
% \documentclass[useAMS,usenatbib,referee]{biom}
% \usepackage{amssymb, amsthm, amsmath}
\usepackage{amssymb, amsmath}
\usepackage{bm}
\usepackage{graphicx}
% \usepackage[authoryear]{natbib}
\usepackage{natbib}
\usepackage{bm}
\usepackage{verbatim}
\usepackage{lineno}
\usepackage{times}
\usepackage{soul}
\usepackage{color}
\usepackage{enumitem}
% \usepackage{enumerate}
\usepackage{setspace}

@@

\usepackage[left=1in,top=1in,right=1in]{geometry}
\pdfpageheight 11in
\pdfpagewidth 8.5in
\linespread{2.0}
\input{../mycommands.sty}

\begin{document} %\linenumbers

% \title[A Space-time Skew-$t$ Model for Threshold Exceedances]{A space-time skew-$t$ model for threshold exceedances}
% \author
% {Samuel A Morris$^{1,*}$\email{samorris@ncsu.edu},
% Brian J Reich$^{1}$,
% Emeric Thibaud$^{2}$, and
% Daniel Cooley$^{2}$\\
% $^{1}$Department of Statistics, North Carolina State University, Raleigh, North Carolina, U.S.A. \\
% $^{2}$Department of Statistics, Colorado State University, Fort Collins, Colorado, U.S.A.}

\begin{center}
{\Large {\bf A space-time skew-$t$ model for threshold exceedances}}\\

{\large Samuel A Morris\footnote[1]{North Carolina State University}, Brian J Reich\footnotemark[1]{}, Emeric Thibaud\footnote[2]{Colorado State University}, and Daniel Cooley\footnotemark[2]{}}

\today
\end{center}

@@

\begin{abstract}
To assess the compliance of air quality regulations, the Environmental Protection Agency (EPA) must know if a site exceeds a pre-specified threshold.
In the case of ozone, the threshold for compliance is fixed at 75 parts per billion, which is high, but not extreme at all locations.
We present a new method based on the spatial skew-$t$ process.
Our method incorporates a random partition to permit long-distance asymptotic independence while allowing for sites that are near one another to be asymptotically dependent, and we incorporate thresholding to allow the tails of the data to speak for themselves.
We also introduce a transformed AR(1) time-series to allow for temporal dependence.
Finally, our model allows for high-dimensional Bayesian inference that is comparable in speed to traditional geostatistical methods for large datasets.
We apply our method to an ozone analysis for July 2005, and find that our model improves over both Gaussian and max-stable methods in terms of predicting exceedances over a fixed threshold.

Keywords: Extreme value analysis, MCMC, Random partition, Skew-$t$, Spatio-temporal
\end{abstract}

% \maketitle

@@

\begin{figure}[h!tbp]
  \centering
  \includegraphics[width=0.75\linewidth]{../plots/ozone-10jul-us.pdf}
  \caption{Ozone values (ppb) on July 10, 2005}
  \label{fig:ozone-10jul}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=0.5\linewidth]{../plots/chi-h.pdf}
  \caption{Extremal dependence measure $\chi(h)$, as a function of distance, $h$, for $K = 1, 3, 5$, and $10$ knots.}
  \label{fig:chi}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{../plots/bsplots-mean.pdf}
  \caption{Brier scores relative to the Gaussian method for simulation study results. A ratio lower than 1 indicates that the method outperforms the Gaussian method.}
  \label{fig:simbrierscores}
\end{figure}

\section{Data analysis}\label{s:analysis}
We consider daily observations of maximum 8-hour ozone measurements for the 31 days of July 2005 at 1,089 Air Quality System (AQS) monitoring sites in the United States as the response (see Figure \ref{fig:ozone-10jul}).
For each site, we also have covariate information containing the estimated ozone from the Community Multi-scale Air Quality (CMAQ) modeling system.
Initially, we fit a linear regression assuming a mean function of
\begin{align}
  \text{E}[Y_i(\bs)] = \beta_0 + \beta_1 \cdot \text{CMAQ}_t(\bs). \label{eq:datamean}
\end{align}
Figure \ref{fig:ozone-qq} shows a Q-Q plot of the residuals compared to a skew-$t$ distribution with 10 d.f. and $\alpha = 1$.
\begin{figure}
  \centering
  \includegraphics[width=0.5\linewidth]{../plots/qq-res.pdf}
  \caption{Q-Q plot of the residuals for a skew-$t$ distribution with 10 d.f. and $\alpha = 1$ (right)}
  \label{fig:ozone-qq}
\end{figure}

Standard exploratory data analysis techniques for extremal dependence are very challenging with only 31 days worth of data because it is difficult to estimate extreme quantiles at each site to obtain empirical estimates of $\chi$.
Despite the fact that there is only one month of data, we can get some sense of extremal dependence between sites by looking at joint occurrences of high sample quantiles.
For example, Figure \ref{fig:bivariateozone} suggests there is more agreement between sites that are close to one another than sites that are far from one another.
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{../plots/daily-quantiles-ozone.pdf}
  \caption{Daily quantiles for two monitoring locations near Columbus, OH (left) and daily quantiles for a monitoring location in Los Angeles, CA and Columbus, OH (right)}
  \label{fig:bivariateozone}
\end{figure}
Another aspect that distinguishes our approach from more traditional extremes analyses, is how the threshold is selected.
In our example, a threshold of 75 ppb which corresponds to $q(92)$ for all observations, but marginally it represents anywhere from $q(0.06)$ to $q(1)$.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{../plots/bs-ozone.pdf}
  \caption{Relative Brier scores for time-series models (left) and non-time-series models (right). Relative brier score for the max-stable model is between 1.07 and 1.15}
  \label{fig:bs-ozone}
\end{figure}
\begin{table}
  \small
  \caption{Top two performing models for ozone analysis at extreme quantiles with Relative Brier score}
  \label{tbl:ozoneresults}
  \centering
  \begin{tabular}{|l|l l l c|l l l c|}
    \cline{2-9}
    \multicolumn{1}{c|}{}  & \multicolumn{4}{c|}{1st} & \multicolumn{4}{c|}{2nd} \\
    \hline
    $q(0.90)$  & No time series & $K=7$  & $T=0$  & BS: 0.980 &
                 No time series & $K=9$  & $T=0$  & BS: 0.980 \\
    $q(0.95)$  & No time series & $K=15$ & $T=50$ & BS: 0.970 &
                 No time series & $K=9$  & $T=50$ & BS: 0.970\\
    $q(0.98)$  & No time series & $K=5$  & $T=50$ & BS: 0.945 &
                 No time series & $K=10$ & $T=50$ & BS: 0.946\\
    $q(0.99)$  & Time series    & $K=10$ & $T=75$ & BS: 0.912 &
                 Time series    & $K=6$  & $T=75$ & BS: 0.913\\
    $q(0.995)$ & Time series    & $K=6$  & $T=75$ & BS: 0.917 &
                 Time series    & $K=10$ & $T=75$ & BS: 0.918\\
    \hline
  \end{tabular}
\end{table}

\begin{figure}
  \centering
  \includegraphics[width=0.8\linewidth]{../plots/q99-ozone.pdf}
  \caption{Panels (a) -- (d) give the posterior predictive $\widehat{q}(0.99)$ for the month of July under four different models, panel (e) gives the difference between $\widehat{q}(0.99)$ in panels (d) and (a), panel (f) gives the difference between $\widehat{q}(0.99)$ in panels (d) and (b).}
  \label{fig:ozoneq99}
\end{figure}

@@

% \backmatter

@@

% \section*{Supplementary Materials}
% Web Appendices A--E are available with this paper at the Biometrics website on Wiley Online Library. \vspace*{-8pt}

\bibliography{library}
\bibliographystyle{rss}
